---
title: "HRT841: Remove influence of surrogate variables from expression matrix"
author: "Miles Roberts"
date: "2021-02-17"
output: html_document
---

## Setup datasets for SVA
```{r}
rm(list = ls())
setwd("C:\\Users\\miles\\OneDrive\\Education\\MichiganStateUniversity\\classes\\fall_2020\\hrt841\\final_project\\05_data_exploration")

library(sva)	#For surrogate variable estimation
library(bladderbatch)	#Contains example data
library(limma)	#For differential gene expression, correction of expression data with surrogate variables
library(ggplot2) #For plotting
library(SRAdb)  #To extract bioproject numbers using SRA numbers
library(reshape)
keyfactors = read.csv("factors_v1.csv", header = T)
express = read.csv("RNAseq_metadata_11-30-20.csv", header = T)

express[1:5,1:5]
head(keyfactors)

#Subset out the SRAs with expression values
keyfacsub = keyfactors[which(keyfactors$sra %in% colnames(express)),]
head(keyfacsub)

#Are all SRAs in matrix also in factor sheet?
all(colnames(express)[-1] %in% keyfactors$sra)

#Which one is missing? There's an extra tab in one sample name
colnames(express)[which(!(colnames(express) %in% keyfactors$sra))]
keyfactors$sra[grepl("SRR122139\t", keyfactors$sra)] = "SRR122139"

#Try subsetting again
#Subset out the SRAs with expression values
keyfacsub = keyfactors[which(keyfactors$sra %in% colnames(express)),]
head(keyfacsub)

#Are all SRAs in matrix also in factor sheet? SHOULD GET TRUE THIS TIME
all(colnames(express)[-1] %in% keyfactors$sra)
```

## Before doing batch correction or SVA, do PCA
```{r}
#Change row names of factor sheet to match column names of expression matrix
rownames(keyfacsub) = keyfacsub$sra
head(keyfacsub)

#Subset phenotype data to relevant factors, add sample column for null model
mypheno = keyfacsub[,c("family", "species", "tissue", "stress", "sra")]
head(mypheno)

#expression data must be in matrix, columns are samples, rows are genes or orthogroups in this case
rownames(express) = express[,1]
express = as.matrix(express[,-1])

noCorPca = prcomp(t(express), center = T, scale = T)
plot(noCorPca$x)

#Plot PCA and color points by family, tissue, and stress
plotdata = data.frame(SRAnumber = rownames(noCorPca$x), PC1 = noCorPca$x[,"PC1"], PC2 = noCorPca$x[,"PC2"])

plotfactors = mypheno
plotfactors$SRAnumber = rownames(mypheno)

plotdata = merge(plotdata,plotfactors, by = "SRAnumber")
head(plotdata)

ggplot(data = plotdata, aes(x = PC1, y = PC2, col = family)) +
	theme_classic() +
	geom_point()

ggplot(data = plotdata, aes(x = PC1, y = PC2, col = tissue)) +
	theme_classic() +
	geom_point()

ggplot(data = plotdata, aes(x = PC1, y = PC2, col = stress)) +
	theme_classic() +
	geom_point()
```

## DO ANOVA BEFORE SVA
```{r echo = T, results = 'hide'}
#Before applying SVA, see if interactions between primary variables are significant
#Write out names for output files, if names files already exist, delete them so that append function later doesn't mess with data structure
pvaluesFile = "anovaKruskalPvaluesPreSVA_2021-02-05.txt"
coeffsFile = "anovaCoeffsPreSVA_2021-02-05.txt"

if(file.exists(pvaluesFile)){
  file.remove(pvaluesFile)
}

if(file.exists(coeffsFile)){
  file.remove(coeffsFile)
}

#Loop over each orthogroup, apply ANOV to each orthogroup
for(i in 1:nrow(express)){
  #Print progress as for loop goes
  if(i %% 100 == 0){
    print(i)
  }
  
  #Extract data for just one orthogroup
  expSub = express[i,]
  expSub = data.frame(sra = names(expSub), expression = expSub)
  lmdat = merge(mypheno, expSub, by = "sra")
  
  #Apply anova and non-parametric equivalent
  lmmod = aov(expression ~ family*tissue*stress, data = lmdat)
  kfammod = kruskal.test(expression ~ family, data = lmdat)
  ktismod = kruskal.test(expression ~ tissue, data = lmdat)
  kstrmod = kruskal.test(expression ~ stress, data = lmdat)
  
  #Save coefficients
  #Orthogroup Coefficient Coefficient.value p.value
  newCoeff = lmmod$coefficients
  coeffs = data.frame(orthogroup = rep(rownames(express)[i], times = length(newCoeff)), coefficient = names(newCoeff), coefficient.value = newCoeff)
  
  #Save pvalues
  lmmod = anova(lmmod)
  aovPvalue = data.frame(orthogroup = rep(rownames(express)[i], times = 7), terms = rownames(lmmod)[-8], aovStat = lmmod[-8,4], aovPvalue = lmmod[-8,5])
  
  kruskalPvalue = data.frame(orthogroup = rep(rownames(express)[i], times = 3), terms = c("family", "tissue", "stress"), kruskalStat = c(kfammod$statistic, ktismod$statistic, kstrmod$statistic), krsukalPvalue = c(kfammod$p.value, ktismod$p.value, kstrmod$p.value))
  
  pvalues = merge(aovPvalue, kruskalPvalue, by = c("orthogroup", "terms"), all.x = T)
  
  #Write results to tables
  write.table(pvalues, pvaluesFile, sep = "\t", quote = F, row.names = F, append = T, col.names = F)
  write.table(coeffs, coeffsFile, sep = "\t", quote = F, row.names = F, append = T, col.names = F)
}
```

## Visualize p-values from pre-SVA ANOVA 
```{r}
savedPvalues = read.table(pvaluesFile, sep = "\t")

#Subset significant terms after bonferonni correction
sigPvalues = savedPvalues[which(savedPvalues$V4 < 0.05/nrow(savedPvalues)),]

#histogram frequency of significant values in different levels
hist(sigPvalues[which(sigPvalues$V2 == "family"),"V4"])
hist(sigPvalues[which(sigPvalues$V2 == "tissue"),"V4"])
hist(sigPvalues[which(sigPvalues$V2 == "stress"),"V4"])

#barplot frequency of significant values in different levels
table(sigPvalues$V2)

plotdata = table(sigPvalues$V2)[c("tissue", "family", "stress", "family:tissue", "family:stress", "tissue:stress", "family:tissue:stress")]

par(mar = c(10,6,3,1))
barplot(plotdata, las = 2, ylab = "Frequency of significant ANOVA term")
```

## Apply SVA
```{r}
#Try another model. Specify two-step method because it doesn't require null model specification, which causes issues with unbalanced datasets
mod1 = model.matrix(~family + tissue + stress, data = mypheno)
mod0 =  model.matrix(~1, data = mypheno)
svseq = svaseq(express, mod1, method = "two-step")
plot(svseq$sv, pch=19, col="blue")

#Now correct for surrogate variables, fit linear model then use modeled expression values
modSv = cbind(mod1,svseq$sv)
fit = lmFit(express,modSv)
#attributes(fit)

#Calcualate p-value for differential expression, adjusting for multiple testing
mod0Sv = cbind(mod0, svseq$sv)
pValues = f.pvalue(express,modSv,mod0Sv)
qValues = p.adjust(pValues,method="BH")

#Which orthogroups are NOT differentially expressed? These could be very interesting to look at. Their expression could be conserved across all three primary variables
putHouseGenes = names(qValues[which(qValues >= 0.05)])
putHouseGenes

#Do t-test for coefficients, H0: coefficient == 0, H1: coefficient != 0 (two-tailed test)
#Divide coefficients by standard deviation, take absolute value so that all statistics fall in right tail, then multiply area under curve to right of statistics by 2 (two-tailed test)
svaCoeffSig = pt(abs(fit$coefficients / fit$stdev.unscaled), df = unique(fit$df.residual), lower.tail = F) * 2

#Reformat surrogate variables to include sra numbers
svs = as.data.frame(mod0Sv)
svs[,"(Intercept)"] = rownames(svs)
colnames(svs)[1] = "sra"

#Save results of linear model fit and surrogate variables
write.table(fit$coefficients, "postSVAcoeffs_2021-02-05.txt", quote = F, sep = "\t")
write.table(svaCoeffSig, "postSVAcoeffsRawPvalues_2021-02-05.txt", quote = F, sep = "\t")
write.table(svs, "surrogateVariables_2021-02-05.txt", quote = F, sep = "\t", row.names = F)
```

## Visualize p-values from post-SVA ANOVA
```{r}
postPvalues = read.table("postSVAcoeffsRawPvalues_2021-02-05.txt", sep = "\t")

#Function to count number of significant terms from post-SVA linear model
countSig = function(x){
  y = p.adjust(x, method = "bonferroni")
  return(length(y[which(y < 0.05)]))
}

#Apply function, prepare to make plot
postSigPvalues = apply(postPvalues, MARGIN = 2, FUN = "countSig")

plotdata = data.frame(coefficient = names(postSigPvalues), frequency = postSigPvalues)

#Re-name surrogate variables to match heatmap names
svIndices = which(grepl("X", plotdata$coefficient) & grepl("X.Intercept", plotdata$coefficient) == F)
plotdata$coefficient[svIndices] = gsub("^", "V", as.character(1:length(svIndices)))
plotdata$coefficient[1] = "Intercept"

plotdata$coefficient = as.factor(plotdata$coefficient)

#Reorder barplot bars to look like pareto plot
plotdata$coefficient = reorder(plotdata$coefficient,-plotdata$frequency)

# Basic barplot
ggplot(data = plotdata, aes(x=coefficient, y=frequency)) +
  geom_bar(stat="identity") +
  theme_classic() +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + 
  ylab("Frequency of significant linear model term") +
  xlab("Linear model term")
```

## Regress surrogate variables out of expression matrix and visualize result
```{r}
#Biostars post: https://www.biostars.org/p/121489/
#Define function to remove surrogate variables
cleanY = function(y, mod, svs) {
    X = cbind(mod, svs)
    Hat = solve(t(X) %*% X) %*% t(X)
    beta = (Hat %*% t(y))
    rm(Hat)
    gc()
    P = ncol(mod)
    return(y - t(as.matrix(X[,-c(1:P)]) %*% beta[-c(1:P),]))
}

#Remove surrogate variable effects
cleanExpress = cleanY(express, mod1, svseq$sv)

#Save resulting matrix
write.table(cleanExpress, "expressionMatrixSVAcorrected_2021-02-04.txt", sep = "\t", quote = F)

#Perform PCA on modeled gene expression values data
SraCorPca2 = prcomp(t(cleanExpress), center = T, scale = T)
#plot(SraCorPca2$x)

#Plot PCA and color points by family, tissue, and stress
plotdata = data.frame(SRAnumber = rownames(SraCorPca2$x), PC1 = SraCorPca2$x[,"PC1"], PC2 = SraCorPca2$x[,"PC2"])

plotfactors = mypheno
plotfactors$SRAnumber = rownames(mypheno)

plotdata = merge(plotdata,plotfactors, by = "SRAnumber")
head(plotdata)

ggplot(data = plotdata, aes(x = PC1, y = PC2, col = family)) +
	theme_classic() +
	geom_point()

ggplot(data = plotdata, aes(x = PC1, y = PC2, col = tissue)) +
	theme_classic() +
	geom_point()

ggplot(data = plotdata, aes(x = PC1, y = PC2, col = stress)) +
	theme_classic() +
	geom_point()

```

## Check GO terms in particular orthogroups
```{r}
goterms = read.table("ATH_GO_GOSLIM.txt", skip = 4, header = F, sep = "\t", fill = TRUE )
ogGenes = read.table("Orthogroups_filtered.tsv", header = T, sep = "\t")

putHouseOgs = ogGenes[which(ogGenes$Orthogroup %in% putHouseGenes),]
for(i in 1:nrow(putHouseOgs)){
  araGenes = putHouseOgs[i,"Arabidopsis_thaliana.TAIR10.pep.all"]
  araGenes = strsplit(araGenes, ", ")[[1]]
  araGenes = unique(gsub("\\.[0-9]", "", araGenes))
  
  print(putHouseOgs[i,"Orthogroup"])
  print(unique(goterms[which(goterms$V1 %in% araGenes & goterms$V8 == "C"), c("V1", "V5","V6")]))
}
```

## Correlate surrogate variables with unmodeled technical variables
```{r}
#library(Compositional)
#Load bioproject listings
biop = read.table("bioproject_ids.txt", sep = ",")

biop$V2 = gsub("\\t.*", "", biop$V2)
biop$V2 = gsub(" ", "", biop$V2)
mypheno$sra = rownames(mypheno)
names(biop) = c("sra", "bioproject")
mypheno = merge(mypheno, biop, by = "sra")
rownames(mypheno) = mypheno$sra

#Correlate bioproject with surrogate variables
svBio = merge(mypheno, svs, by = "sra")
str(svBio)

#fit multiple linear regression model
mod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15,V16,V17) ~ bioproject, data = svBio)

#Calculate unadjusted R-squared as 1 - var(residuals)/var(total), then caluclate adjusted R-squared from that because factors with more levels will have more parameters associated with them
extractRsquared = function(mod){
  resvar = sapply(as.data.frame(mod$residuals), var)  #Calculate residual (i.e. unexplained) variation
  totvar = sapply(as.data.frame(mod$residuals + mod$fitted.values), var)  #Calculate total variation in response
  svRsquared = 1 - (resvar/totvar)  #Calcuate R-squared
  k = nrow(mod$coefficients) - 1  #number of predictors
  n = nrow(mod$fitted.values) #Number of observations
  svAdjRsquared = 1 - ((1-svRsquared)*(n-1))/(n-k-1)  #Calculate adjusted R-squared
  return(svAdjRsquared)
}

#compare R-squared values with model output
#summary(mod)
#extractRsquared(mod)

#fit multiple linear regression model
  #Technical variable as predictor
bioMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15,V16,V17) ~ bioproject, data = svBio)
speMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15,V16,V17) ~ species, data = svBio)
  #Primary variable as predictor
famMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15,V16,V17) ~ family, data = svBio)
tisMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15,V16,V17) ~ tissue, data = svBio)
strMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15,V16,V17) ~ stress, data = svBio)

#Extract R-squared values
plotdata = as.data.frame(rbind(extractRsquared(bioMod), extractRsquared(speMod), extractRsquared(famMod), extractRsquared(tisMod), extractRsquared(strMod)))
names(plotdata) = gsub("^", "V", as.character(1:length(names(plotdata)))) #Change surrogate variable names
plotdata = cbind(c("bioproject", "species", "family", "tissue", "stress"),plotdata)
plotdata = melt(plotdata)
names(plotdata) = c("Technical.Variable", "Surrogate.Variable", "Adj.R.squared")

#Plot r-squared values on heatmap
ggplot(plotdata, aes(Surrogate.Variable, Technical.Variable)) +
  theme_minimal() +
  scale_fill_gradient(low = "black", high = "deepskyblue", name = "Adjusted R-squared")   +
  geom_tile(aes(fill = Adj.R.squared)) +
  xlab("Surrogate variable ID") +
  ylab("Primary or unmodeled variable") + 
  theme(text = element_text(size = 12))
```

## Re-run analysis after removing outliers based on rPCA results
### Remove outliers
```{r}
#The below samples look like outliers based on rPCA
outliers = c("SRR089703", "ERR3814227", "ERR392057", "SRR4048280")

#Remove outlier samples from expression matrix, factor sheet
expNoOut = express[,which(!(colnames(express) %in% outliers))]
dim(express)
dim(expNoOut)

phenoNoOut = mypheno[which(!(mypheno$sra %in% outliers)),]
dim(mypheno)
dim(phenoNoOut)
```
### Pre-SVA ANOVA
```{r echo = T, results = 'hide'}
#Before applying SVA, see if interactions between primary variables are significant
#Write out names for output files, if names files already exist, delete them so that append function later doesn't mess with data structure
pvaluesFile = "anovaKruskalPvaluesPreSVA_NoOutliers_2021-02-16.txt"
coeffsFile = "anovaCoeffsPreSVA_NoOutliers_2021-02-16.txt"

if(file.exists(pvaluesFile)){
  file.remove(pvaluesFile)
  write.table(c("orthogroup  terms  aovStat  aovPvalue  kruskalStat  krsukalPvalue"), pvaluesFile, sep = "\t", quote = F, row.names = F, append = T, col.names = F)
}
  
if(file.exists(coeffsFile)){
  file.remove(coeffsFile)
  write.table(c("orthogroup  coefficient  coefficient.value"), coeffsFile, sep = "\t", quote = F, row.names = F, append = T, col.names = F)
}

#Loop over each orthogroup, apply ANOV to each orthogroup
for(i in 1:nrow(expNoOut)){
  #Print progress as for loop goes
  if(i %% 10 == 0){
    print(i)
  }
  
  #Extract data for just one orthogroup
  expSub = expNoOut[i,]
  expSub = data.frame(sra = names(expSub), expression = expSub)
  lmdat = merge(phenoNoOut, expSub, by = "sra")
  
  #Apply anova and non-parametric equivalent
  lmmod = aov(expression ~ family*tissue*stress, data = lmdat)
  kfammod = kruskal.test(expression ~ family, data = lmdat)
  ktismod = kruskal.test(expression ~ tissue, data = lmdat)
  kstrmod = kruskal.test(expression ~ stress, data = lmdat)
  
  #Save coefficients
  #Orthogroup Coefficient Coefficient.value p.value
  newCoeff = lmmod$coefficients
  coeffs = data.frame(orthogroup = rep(rownames(expNoOut)[i], times = length(newCoeff)), coefficient = names(newCoeff), coefficient.value = newCoeff)
  
  #Save pvalues
  lmmod = anova(lmmod)
  aovPvalue = data.frame(orthogroup = rep(rownames(expNoOut)[i], times = 7), terms = rownames(lmmod)[-8], aovStat = lmmod[-8,4], aovPvalue = lmmod[-8,5])
  
  kruskalPvalue = data.frame(orthogroup = rep(rownames(expNoOut)[i], times = 3), terms = c("family", "tissue", "stress"), kruskalStat = c(kfammod$statistic, ktismod$statistic, kstrmod$statistic), krsukalPvalue = c(kfammod$p.value, ktismod$p.value, kstrmod$p.value))
  
  pvalues = merge(aovPvalue, kruskalPvalue, by = c("orthogroup", "terms"), all.x = T)
  
  #Write results to tables
  write.table(pvalues, pvaluesFile, sep = "\t", quote = F, row.names = F, append = T, col.names = F)
  write.table(coeffs, coeffsFile, sep = "\t", quote =  F, row.names = F, append = T, col.names = F)
}
```

### Visualize Pre-SVA results
```{r}
savedPvalues = read.table(pvaluesFile, sep = "\t", skip = 1)

#Subset significant terms after bonferonni correction
sigPvalues = savedPvalues[which(savedPvalues$V4 < 0.05/nrow(savedPvalues)),]

#histogram frequency of significant values in different levels
hist(sigPvalues[which(sigPvalues$V2 == "family"),"V4"])
hist(sigPvalues[which(sigPvalues$V2 == "tissue"),"V4"])
hist(sigPvalues[which(sigPvalues$V2 == "stress"),"V4"])

#barplot frequency of significant values in different levels
table(sigPvalues$V2)

plotdata = table(sigPvalues$V2)[c("tissue", "family", "stress", "family:tissue", "family:stress", "tissue:stress", "family:tissue:stress")]

par(mar = c(10,6,3,1))
barplot(plotdata, las = 2, ylab = "Frequency of significant ANOVA term")
```

### Apply SVA after outliers are removed
```{r}
#Specify two-step method because it doesn't require null model specification, which causes issues with unbalanced datasets
mod1 = model.matrix(~family + tissue + stress, data = phenoNoOut)
mod0 =  model.matrix(~1, data = phenoNoOut)
svseq = svaseq(expNoOut, mod1, method = "two-step")
plot(svseq$sv, pch=19, col="blue")

#Now correct for surrogate variables, fit linear model then use modeled expression values
modSv = cbind(mod1,svseq$sv)
fit = lmFit(expNoOut,modSv)
#attributes(fit)

#Calcualate p-value for differential expression, adjusting for multiple testing
mod0Sv = cbind(mod0, svseq$sv)
pValues = f.pvalue(expNoOut,modSv,mod0Sv)
qValues = p.adjust(pValues,method="BH")

#Which orthogroups are NOT differentially expressed? These could be very interesting to look at. Their expression could be conserved across all three primary variables
putHouseGenes = names(qValues[which(qValues >= 0.05)])
putHouseGenes

#Do t-test for coefficients, H0: coefficient == 0, H1: coefficient != 0 (two-tailed test)
#Divide coefficients by standard deviation, take absolute value so that all statistics fall in right tail, then multiply area under curve to right of statistics by 2 (two-tailed test)
svaCoeffSig = pt(abs(fit$coefficients / fit$stdev.unscaled), df = unique(fit$df.residual), lower.tail = F) * 2

#Reformat surrogate variables to include sra numbers
svs = as.data.frame(mod0Sv)
svs[,"(Intercept)"] = rownames(svs)
colnames(svs)[1] = "sra"

#Save results of linear model fit and surrogate variables
write.table(fit$coefficients, "postSVAcoeffs_NoOutliers_2021-02-16.txt", quote = F, sep = "\t")
write.table(svaCoeffSig, "postSVAcoeffsRawPvalues_NoOutliers_2021-02-16.txt", quote = F, sep = "\t")
write.table(svs, "surrogateVariables_NoOutliers_2021-02-16.txt", quote = F, sep = "\t", row.names = F)
```

### Post-SVA ANOVA, outliers removed before SVA
```{r}
postPvalues = read.table("postSVAcoeffsRawPvalues_NoOutliers_2021-02-16.txt", sep = "\t")

countSig = function(x){
  y = p.adjust(x, method = "bonferroni")
  return(length(y[which(y < 0.05)]))
}

postSigPvalues = apply(postPvalues, MARGIN = 2, FUN = "countSig")

plotdata = data.frame(coefficient = names(postSigPvalues), frequency = postSigPvalues)

svIndices = which(grepl("X", plotdata$coefficient) & grepl("X.Intercept", plotdata$coefficient) == F)
plotdata$coefficient[svIndices] = gsub("^", "V", as.character(1:length(svIndices)))
plotdata$coefficient[1] = "Intercept"

plotdata$coefficient = as.factor(plotdata$coefficient)

plotdata$coefficient = reorder(plotdata$coefficient,-plotdata$frequency)

# Basic barplot
ggplot(data = plotdata, aes(x=coefficient, y=frequency)) +
  geom_bar(stat="identity") +
  theme_classic() +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + 
  ylab("Frequency of significant linear model term") +
  xlab("Linear model term")
```

### Correlate surrogate variables with unmodeled technical variables
```{r}
#library(Compositional)
#Load bioproject listings
biop = read.table("bioproject_ids.txt", sep = ",")

biop$V2 = gsub("\\t.*", "", biop$V2)
biop$V2 = gsub(" ", "", biop$V2)
phenoNoOut$sra = rownames(phenoNoOut)
names(biop) = c("sra", "bioproject")
phenoNoOut = merge(phenoNoOut, biop, by = "sra")
rownames(phenoNoOut) = phenoNoOut$sra

#Correlate bioproject with surrogate variables
svBio = merge(phenoNoOut, svs, by = "sra")
str(svBio)

#fit multiple linear regression model
mod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15) ~ bioproject, data = svBio)

#Calculate unadjusted R-squared as 1 - var(residuals)/var(total), then caluclate adjusted R-squared from that because factors with more levels will have more parameters associated with them
extractRsquared = function(mod){
  resvar = sapply(as.data.frame(mod$residuals), var)  #Calculate residual (i.e. unexplained) variation
  totvar = sapply(as.data.frame(mod$residuals + mod$fitted.values), var)  #Calculate total variation in response
  svRsquared = 1 - (resvar/totvar)  #Calcuate R-squared
  k = nrow(mod$coefficients) - 1  #number of predictors
  n = nrow(mod$fitted.values) #Number of observations
  svAdjRsquared = 1 - ((1-svRsquared)*(n-1))/(n-k-1)  #Calculate adjusted R-squared
  return(svAdjRsquared)
}

#compare R-squared values with model output
#summary(mod)
#extractRsquared(mod)

#fit multiple linear regression model
  #Technical variable as predictor
bioMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15) ~ bioproject, data = svBio)
speMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15) ~ species, data = svBio)
  #Primary variable as predictor
famMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15) ~ family, data = svBio)
tisMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15) ~ tissue, data = svBio)
strMod = lm(cbind(V2,V3,V4,V5,V6,V7,V8,V9,V10,V11,V12,V13,V14,V15) ~ stress, data = svBio)

#Extract R-squared values
plotdata = as.data.frame(rbind(extractRsquared(bioMod), extractRsquared(speMod), extractRsquared(famMod), extractRsquared(tisMod), extractRsquared(strMod)))
names(plotdata) = gsub("^", "V", as.character(1:length(names(plotdata)))) #Change surrogate variable names
plotdata = cbind(c("bioproject", "species", "family", "tissue", "stress"),plotdata)
plotdata = melt(plotdata)
names(plotdata) = c("Technical.Variable", "Surrogate.Variable", "Adj.R.squared")

#Plot r-squared values on heatmap
ggplot(plotdata, aes(Surrogate.Variable, Technical.Variable)) +
  theme_minimal() +
  scale_fill_gradient(low = "black", high = "deepskyblue", name = "Adjusted R-squared")   +
  geom_tile(aes(fill = Adj.R.squared)) +
  xlab("Surrogate variable ID") +
  ylab("Primary or unmodeled variable") + 
  theme(text = element_text(size = 12))
```